Revisiting the Austral Spring Extratropical Southern Hemisphere zonally asymmetric circulation using complex Empirical Orthogonal Functions

### 2022-06-03 analysis/paper/shceof.Rmd
Agerga comentarios Leandro

We analyse the joint variabilty of the zonally asymmetric springtime stratospheric and tropospheric circulation using Complex Empirical Orthogonal Functions (cEOF) to characterise planetary waves of varying amplitude and phase. 
The leading cEOF represents variabilty of a zonal wave 1 troposphere that correlates slightly with the Symmetric Southern Annular Mode (S-SAM).
The second cEOF (cEOF2) is an alternative representation of the Pacific-South American modes. 
One phase of this cEOF is also extremely correlated with the Asymmetric SAM (A-SAM) in the troposphere. 
Springs with an active ENSO tend to phase-lock the cEOF2, but have no consistent impact on its magnitude.
Furthermore, we find indications that the location of Pacific Sea Surface Temperature anomalies affect the phase of the cEOF2.

@@

Both the spatial pattern similarity and the high temporal correlation of cOEFs computed at 50 hPa and 200 hPa suggest that these are, to a large extent, modes of joint variability.
This motivates the decision of performing complex EOF jointly between levels.
Therefore cEOFs were computed using data from both levels at the same time.
In that sense each cEOF has a spatial component that depends on longitude, latitude and level, and a temporal component that depends only on time.

@@

Figure \@ref(fig:eof1-regr-gh) shows regression maps of SON geopotential height anomalies upon cEOF1.
At 50 hPa (Figure \@ref(fig:eof1-regr-gh) row a), the Real cEOF1 is associated with a centre located over the Ross Sea.
The Imaginary cEOF1 is associated with a distinctive wave 1 pattern with maximum over the coast of East Antarctica.

At 200 hPa (Figure \@ref(fig:eof1-regr-gh) row b) the Real cEOF1 shows a single centre of positive anomalies spanning West Antarctica surrounded by opposite anomalies in lower latitudes, with is centre shifted slightly eastward compared with the upper-level anomalies.
The Imaginary cEOF1 shows a much more zonally symmetrical pattern resembling the negative SAM phase [e.g. @fogt2020].

Therefore, the magnitude and phase of the cEOF1 are associated with the magnitude and phase of a zonal wave mainly in the stratosphere.

@@

Therefore, by virtue of being the most common phase, the Imaginary cEOF2 explains more variance than the Real cEOF2.
Conventional EOF analysis will therefore tend to separate them relatively cleanly, with the EOF representing the Imaginary cEOF2 always leading the one representing the Real cEOF2.

This phase preferences is in agreement with @irving2016, who found a bimodal distribution to PSA-like variability (compare our Figure \@ref(fig:phase-histogram) with their Figure 6).

@@

In the case of Imaginary cEOF2, the correlation between SAM and Imaginary cEOF2 for the troposphere is associated to the asymmetric variability of the SAM. 
Indeed, the Imaginary cEOF2 shares up to `r scales::percent(max_r2[cEOF == "cEOF2" & term == "asym"]$estimate^2, 2)` variance with the A-SAM and only `r scales::percent(max_r2[cEOF == "cEOF2" & term == "sym"]$estimate^2, 2)` at most with the S-SAM (Figure \@ref(fig:sam-eof-vertical).b2).
Such extremely high correlation between A-SAM and Imaginary cEOF2 suggests that the modes obtained in this work are able to characterise the zonally asymmetric component of the SAM described previously by @campitelli2022.

@@

The correlation between the absolute magnitude of the ONI index and the  cEOF2 amplitude is `r cor_enso_mag$text`. 
However, this relationship is mostly driven by the three years with strongest ENSO events in the period (`r knitr::combine_words(year(gdata[order(-oni)][1:3]$time))`) which coincide with the three years with strongest cEOF2 magnitude (not shown).
If those years are removed, the correlation becomes non-significant (`r cor_enso_mag_outlier$text`).
Furthermore, even when using all years, the Spearman correlation --which is robust to outliers-- is also non-significant (`r signif(cor_enso_spearman$estimate, 2)`, p-value`r report_p(cor_enso_spearman$p.value)`). 
Therefore, although the location of tropical SST anomalies seem to have an effect in defining the phase of the cEOF2, the relationship between the magnitude of cEOF2 and ONI remains uncertain might be evident only in particularly intense ENSO events, for which there are few data.

@@

The influence of cEOFs variability in the anomalies of both 2-metre air temperature and precipitation in the Southern Hemisphere is also explored.
Figure\ \@ref(fig:pp-t2m-r2) shows the 2-meter temperature or precipitation anomalies explained variance by the multiple linear model of both Real and Imaginary cEOF1 components (column 1), and both Real  and Imaginary cEOF2 components (column 2).
The variance explained by cEOF1 for precipitation anomalies is extremely low and also for most of the temperature anomalies, except for the northern tip of the Antarctic Peninsula, northern Weddell Sea and the Ross Sea coast (Fig.\  \@ref(fig:pp-t2m-r2)a.1).

@@

In previous sections, it has been shown that variability patterns that arise from cEOF methodology describe the zonally asymmetric springtime extratropical SH circulation, reproducing previous features such as the variability related to PSAs or A-SAM.

@@

Our method is similar to @goyal2022 as they also derive an index with amplitude and phase of wave-like activity in the SH with very similar patterns to our cEOF2. 
Both indices are likely very similar (a detailed comparison is out of scope for this paper) but we believe that cEOF have the advantage of constructing the indices based on patterns that are exactly in quadrature by construction.

>>>>

The large-scale extratropical circulation in the Southern Hemisphere is strongly zonally symmetric, but its zonal departures, albeit highly relevant for regional impacts, have been less studied.
In this study we analyse the joint variability of the zonally asymmetric springtime stratospheric and tropospheric circulation using Complex Empirical Orthogonal Functions (cEOF) to characterise planetary waves of varying amplitude and phase.
The leading cEOF represents variability of a zonal wave 1 in the stratosphere that correlates slightly with the Symmetric Southern Annular Mode (S-SAM).
The second cEOF (cEOF2) is an alternative representation of the Pacific-South American modes. 
One phase of this cEOF is also extremely correlated with the Asymmetric SAM (A-SAM) in the troposphere. 
Springs with an active ENSO tend to phase-lock the cEOF2, but have no consistent impact on its magnitude.
Furthermore, we find indications that the location of Pacific Sea Surface Temperature anomalies affect the phase of the cEOF2. 
As a result, the methodology proposed in this study provides a deeper understanding of the zonally asymmetric springtime extratropical SH circulation.

@@

Both the spatial pattern similarity and the high temporal correlation of cEOFs computed at 50 hPa and 200 hPa suggest that these are, to a large extent, modes of joint variability.
This motivates the decision of performing complex EOF jointly between levels.
Therefore cEOFs were computed using data from both levels at the same time.
In that sense each cEOF has a spatial component that depends on longitude, latitude and level, and a temporal component that depends only on time.

@@

Figure \@ref(fig:eof1-regr-gh) shows regression maps of SON geopotential height anomalies upon cEOF1.
At 50 hPa (Figure \@ref(fig:eof1-regr-gh) row a), the Real cEOF1 is associated with a centre located over the Ross Sea.
The Imaginary cEOF1 is associated with a distinctive wave 1 pattern with maximum over the coast of East Antarctica. 
At 200 hPa (Figure \@ref(fig:eof1-regr-gh) row b) the Real cEOF1 shows a single centre of positive anomalies spanning West Antarctica surrounded by opposite anomalies in lower latitudes, with is centre shifted slightly eastward compared with the upper-level anomalies.
The Imaginary cEOF1 shows a much more zonally symmetrical pattern resembling the negative SAM phase [e.g. @fogt2020].
Therefore, the magnitude and phase of the cEOF1 are associated with the magnitude and phase of a zonal wave mainly in the stratosphere.

@@

Therefore, by virtue of being the most common phase, the Imaginary cEOF2 explains more variance than the Real cEOF2.
Conventional EOF analysis will therefore tend to separate them relatively cleanly, with the EOF representing the Imaginary cEOF2 always leading the one representing the Real cEOF2.
This phase preferences is in agreement with @irving2016, who found a bimodal distribution to PSA-like variability (compare our Figure \@ref(fig:phase-histogram) with their Figure 6).

@@

In the case of Imaginary cEOF2, its correlation with the SAM for the troposphere is associated to the asymmetric variability of the SAM. 
Indeed, the Imaginary cEOF2 shares up to `r scales::percent(max_r2[cEOF == "cEOF2" & term == "asym"]$estimate^2, 2)` variance with the A-SAM and only `r scales::percent(max_r2[cEOF == "cEOF2" & term == "sym"]$estimate^2, 2)` at most with the S-SAM (Figure \@ref(fig:sam-eof-vertical).b2).
Such extremely high correlation between A-SAM and Imaginary cEOF2 suggests that the modes obtained in this work are able to characterise the zonally asymmetric component of the SAM described previously by @campitelli2022.

@@

The correlation between the absolute magnitude of the ONI index and the  cEOF2 amplitude is `r cor_enso_mag$text`. 
However, this relationship is mostly driven by the three years with strongest ENSO events in the period (`r knitr::combine_words(year(gdata[order(-oni)][1:3]$time))`) which coincide with the three years with strongest cEOF2 magnitude (not shown).
If those years are removed, the correlation becomes non-significant (`r cor_enso_mag_outlier$text`).
Furthermore, even when using all years, the Spearman correlation --which is robust to outliers-- is also non-significant (`r signif(cor_enso_spearman$estimate, 2)`, p-value`r report_p(cor_enso_spearman$p.value)`). 
Therefore, although the location of tropical SST anomalies seem to have an effect in defining the phase of the cEOF2, the relationship between the magnitude of cEOF2 and ONI remains uncertain and might be only evident in very strong ENSO events, that are scarce in the historical observational record.

@@

The influence of cEOFs variability in the anomalies of both 2-metre air temperature and precipitation in the SH was also explored.
Figure\ \@ref(fig:pp-t2m-r2) shows the 2-meter temperature or precipitation anomalies explained variance by the multiple linear model of both Real and Imaginary cEOF1 components (column 1), and both Real  and Imaginary cEOF2 components (column 2).
The variance explained by cEOF1 for precipitation anomalies is extremely low and also for most of the temperature anomalies, except for the northern tip of the Antarctic Peninsula, northern Weddell Sea and the Ross Sea coast (Fig.\  \@ref(fig:pp-t2m-r2)a.1).

@@

In this study we assessed extratropical Southern Hemisphere zonally asymmetric circulation in austral spring. For this purpose, two complex indices were derived using Complex Empirical Orthogonal Functions and used to characterise both amplitude and phase of planetary waves.

The cEOF1 represents the variability of the zonal wave 1 in the stratosphere and is closely related to stratospheric variability such as anomalies in Total Ozone Column. Otherwise, this complex EOF is not related with SST variability and continental precipitation in the Southern Hemisphere. On the other hand, the cEOF2 represents a wave-3 pattern with maximum magnitude in the Pacific sector, that is an alternative representation of the PSA1 and PSA2 patterns (Mo and Paegle 2001). The Imaginary cEOF2 can be identified with the PSA1 and the Real cEOF2 with the PSA2.

While the cEOF2 variability could be related to surface impacts, the cEOF1 influence is almost negligible. For instance, precipitation anomalies in South America associated with the Imaginary cEOF2 show a clear ENSO-like impact, with positive anomalies in South-Eastern South America, negative anomalies in Southern Brazil and positive anomalies in central Chile for positive Imaginary cEOF2 phase.

Variability patterns that arise from cEOF methodology describe the zonally asymmetric springtime extratropical SH circulation, reproducing previous features such as the variability related to PSAs or A-SAM.

@@

Our method is similar to @goyal2022 as they also derive an index with amplitude and phase of wave-like activity in the SH with very similar patterns to our cEOF2. 
Both indices are likely very similar (a detailed comparison is out of scope for this paper) but cEOF have the advantage of constructing the indices based on patterns that are exactly in quadrature by construction.
